---
title: "House Sales Price Prediction Using the Ames Assessor’s Office Data"
author: "Boyko Amarov"
format: pptx
---

```{r setup, echo=FALSE, warning=FALSE}
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse", repos = "http://cran.us.r-project.org")
}

if (!requireNamespace("GGally", quietly = TRUE)) {
  install.packages("GGally", repos = "http://cran.us.r-project.org")
}

if (!requireNamespace("rmarkdown", quietly = TRUE)) {
  install.packages("rmarkdown", repos = "http://cran.us.r-project.org")
}

if (!requireNamespace("broom", quietly = TRUE)) {
  install.packages("broom", repos = "http://cran.us.r-project.org")}
# Load the libraries

library(tidyverse)
library(readr)
library(readxl)
library(GGally)
library(broom)

# You should download _your_ data in this folder and change the name of the file in the code below

dt <- read_csv("AmesHousing.csv")

# Example reading from an Excel file (to run the code you should remove the # character from the beginning of the line below)

# dt <- read_excel("AmesHousing.xlsx")

# Rename the variables that are used in the analysis to avoid names with spaces and special characters

dt <- dt %>% 
    rename(
      # The syntax of rename is: new_name = old_name, if the old name has spaces or special characters, put it in backticks
        LotArea = `Lot Area`,
        YearBuilt = `Year Built`,
        LivingArea = `Gr Liv Area`
    )
```

## Introduction

-   The Ames Assessor’s Office dataset provides detailed information about residential properties in Ames, Iowa between 2006 and 2010. <https://jse.amstat.org/v19n3/decock/DataDocumentation.txt>
-   The dataset contains 2930 observations and 82 variables.
-   In this presentation we explore a linear regression model for the prediction of house sales prices based on various features of the properties.

## Research Question

What is the association between house sales prices and the following features of the properties?

-   Living area (in square feet)
-   Lot area (in square feet)
-   Year built (in years)


## The Data

```{r}
# Create a scatterplot matrix to visualize the relationships between the variables of interest

fig <- dt %>% 
  select(LivingArea, LotArea, YearBuilt, SalePrice) %>% 
  ggpairs(
    title = "Scatterplot Matrix of House Value and Features",
    lower = list(continuous = wrap("points", alpha = 0.2)),
    upper = list(continuous = wrap("cor", size = 5))
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
# Save the plot as a PNG file

ggsave("scatterplot_matrix.png", plot = fig, width = 10, height = 8)

fig
```

## The Model

$$
\begin{align*}
\text{Price}_i & = \beta_0 + \beta_1 \text{LivingArea}_i + \beta_2 \text{LotArea}_i + \beta_3\text{YearBuilt}_i + e_i
\end{align*}
$$

where $e_i$ are zero-mean, uncorrelated random terms with constant variance $\sigma^2$.

- The model is estimated using the method of least squares.
- The predictors are centered by subtracting their mean to improve the interpretability of the coefficients.
- The sales price is divided by 1000 to avoid large coefficients.

## Results

```{r lm-fit}

dt <- dt %>%
  mutate(
    SalesPriceC = SalePrice / 1000,
    LotAreaC = LotArea - mean(LotArea),
    YearBuiltC = YearBuilt - mean(YearBuilt),
    LivingAreaC = LivingArea - mean(LivingArea)
  )

fit <- lm(SalesPriceC ~ LivingAreaC + LotAreaC + YearBuiltC, data = dt)

summary(fit) %>% 
  tidy() %>% 
  select(term, estimate, std.error, p.value) %>%
  knitr::kable(
    col.names = c("Variable", "Estimate", "Std. Error", "p value"),
    digits = 3
  )
```

R squared: `r summary(fit)$r.squared`

## Alternative Output

If you encounter difficulties with the table from the previous slide, just 
run the code below to get the summary of the model and paste it in your PowerPoint presentation.

```{r lm-fit-alt}
summary(fit)
```

## Interpretation (I)

- The estimated sales price for a house with an average year built, average lot area, and average living area is 180,796 USD.
- Newer houses tended to be more expensive. A difference of 10 years is associated with about 10,990 USD price difference (for equal lot area and living area).
- As expected: larger houses tend to be more expensive. A difference of 100 square feet is associated with about 9,200 USD higher price (for equal lot area and year built).

## Interpretation (II)

- Houses with larger lot areas tend to be more expensive. 
- All coefficients are statistically significant at the 0.05 level.


## Goodness of Fit

```{r goodness-of-fit}
# Plot the predicted vs actual values

# If the model fits the data well, the points should be close to the 45-degree line and should not show any systematic pattern other 
# than random scatter around the line.

# For this example the plot shows a poor fit for the higher priced houses as they are systematically under-predicted by the model.
# A solution could be to use a log transformation of the dependent variable or to include non-linear terms in the model.

fig <- 
  dt %>%
  mutate(
    predicted_values = predict(fit)
  ) %>%
  ggplot(aes(x = predicted_values, y = SalesPriceC)) +
  geom_point(alpha = 0.2) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Actual vs. Predicted Sales Prices",
    x = "Predicted Sales Price (in thousands)",
    y = "Actual Sales Price (in thousands)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

ggsave("predicted_vs_actual.png", plot = fig, width = 10, height = 8)

fig
```


## Last Slide

Thank you for your attention!
